---
title: "Class Lab 1"
author: "Raymond Atwebembere"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(haven) 
library(readr) 
library(dplyr) 
library(openxlsx) 

pacman::p_load(haven, readr, dplyr, openxlsx)
```

INSTRUCTIONS Complete all tasks 1-7 using reproducible coding practices (annotation of all code). Use R syntax to solve all problems (i.e. do not manually change values/variables in datasets or check them manually to make sure things are correct except as a double or triple check). Please cut and paste each problem before each R markdown chunk and annotate your code within the chunks. Compare and consolidate your answers. Select one member of your group to walk the class through how each problem was solved. Elect one member of your group to turn in your consolidated code with answers. Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

1. Import class one survey data from our Github site (<https://github.com/kijohnson/ADA-2024>) (The dataset is called ‘Class 1 Survey Fall 2024_di.csv’), calling the R dataframe that you create C1survey.

```{r}

C1survey <-
 read_csv(
   "https://raw.githubusercontent.com/kijohnson/ADA-2024/main/Class%201%20(Getting%20started%20and%20RR%20I)/Lab/Class%201%20Survey%20Fall%202024_di.csv"
   )

```

2.  Determine: a. the number of observations (i.e. the number of people who filled out the survey) and b. the number of variables in the dataframe.

```{r}

dim(C1survey) 
str(C1survey)
```

3.   a.  Rename the column variables to something shorter and that is descriptive of what the variable is about (for example like_dogs for the ‘Do you like dogs?’ question variable) and b. Display that they are renamed.

```{r}
# rename variables in order of appearance

colnames(C1survey)

names(C1survey)[1:27]<-c("id", "like_cats", "like_dogs", "have_desert", "slogan", "fav_day", "larkORowl", "fav_food", "fav_drink", "fav_season", "fav_month", "hobby", "program", "specialization", "stat_software", "R_exp", "coding_comfort", "coding_length", "top_three","public_health_interest", "fav_num", "bday", "bmonth", "country", "state", "city", "highest_educ_level")

colnames(C1survey)

```

4.  Write code to determine and display the number of factor, integer, numerical, and character variables there are in the C1survey dataset.

```{r}
table(sapply(C1survey, class)) #Best solution to use in this case

#or, depending on how you want to see

data_structure <- sapply(C1survey, typeof)
print (data_structure)
```

5.  a.) Using code, check the bday and bmonth variables for any unusual or missing values. If you find any, b.) describe what you will do with the unusual values in a comment before or immediately following the code that does it, and c.) after you have cleaned up any unusual values, find the median bday and month. *Note: please make sure that your bday and month variables are numeric before doing statistics.

```{r}
#a. look at the values
class(C1survey$bday) #C1survey$bday
class(C1survey$bmonth) #C1survey$bmonth

# recoding bday with non numeric values
class(C1survey$bday)
C1survey$bday<-ifelse(C1survey$bday == "May 21-report 21", "21", C1survey$bday)
C1survey$bday<-ifelse(C1survey$bday == "March 31st", "31", C1survey$bday)
C1survey$bday
# convert to numeric for statistics
C1survey$bday<-as.numeric(C1survey$bday)
table(C1survey$bday, useNA = "always")

#b. bmonth recoding method 1:

C1survey$bmonth[C1survey$bmonth=="May 21-report 5"]<-"5"
C1survey$bmonth[C1survey$bmonth=="February"]<-"2"
C1survey$bmonth[C1survey$bmonth=="September"]<-"9"
C1survey$bmonth[C1survey$bmonth=="July"]<-"7"
C1survey$bmonth[C1survey$bmonth=="March"]<-"3"
C1survey$bmonth # check values 

# convert to numeric for statistics
C1survey$bmonth<-as.numeric(C1survey$bmonth)
C1survey$bmonth # check values 

#I have detected an unnsual value, which is zero. This may skew the measures of central tendency.

# c. calculate medians for bday and bmonth 
median(C1survey$bday, na.rm=TRUE) # 14 (biased because of 0, which is not a value) 

# How could we better design a survey to ensure this doesn't happen?

#I would ensure that I put a condition in the data collecting tool not to accept 0 as an entry

median(C1survey$bmonth, na.rm=TRUE) # 7
```

6. a.  Create a new variable called bseason that gives the season according to Northern Meteorological season in which respondents were born (winter = 12/1 to 2/29, spring = 3/1 to 5/31, summer = 6/1 to 8/31, fall = 9/1 to 11/30). b. Using the table function, print a table of Seasons in the columns and bmonths in the rows that allows you to check that the coding is correct c. Sum the columns using the addmargins function to answer the question of how many classmates were born in each season? Include your answer in the code.

```{r}

# Assigning seasons based on birth month
assignSeason <- function(month) {
  if (month %in% c(12, 1, 2)) {
    return("Winter")
  } else if (month %in% c(3, 4, 5)) {
    return("Spring")
  } else if (month %in% c(6, 7, 8)) {
    return("Summer")
  } else if (month %in% c(9, 10, 11)) {
    return("Fall")
  } 
}

#b Applying function to the bmonth column to create the bseason column
C1survey$bseason <- sapply(C1survey$bmonth, assignSeason)

#c Creating a table of Seasons in the columns and bmonths in the rows
season_month_table <- table(C1survey$bmonth, C1survey$bseason)
print(season_month_table)

# Sum the columns using addmargins to find the total number of classmates per season
season_totals <- addmargins(season_month_table, 1)
print(season_totals)

#Sum of classmates born each season include: Fall=2, Spring=7, Summer=7, Winter=4
```

7.  Pick your favorite variable or variables to analyze, come up with a question you want to answer with that variable or variables, generate the code, and provide an answer to your question. Describe what you did using comments (i.e. #’s).

```{r}
###a. Research question: Is there a pet preference among larks, owls and humming birds?

###b Variables considered: i) Pet_preference (like_cats; like_dogs) ii) larkORowl (lark, Owl, Hummingbird)

  #### Creating new categorical variable Preferred_pet:

C1survey <- C1survey %>% 
  mutate(Pet_preference = case_when (
    like_dogs == "Yes" & like_cats == "Yes" ~ "Both",
    like_dogs == "Yes" & like_cats == "No" ~ "Dogs",
    like_dogs == "No" & like_cats == "Yes" ~ "Cats",
                                     TRUE ~ "Neither") )

class(C1survey$Pet_preference) #checking variable type

####Next: create a table exploring both pet_preference and lark_owl_hummingbird 

contingency_table <- table(C1survey$Pet_preference, C1survey$larkORowl)
contingency_table #view the table to know the pet preferences

###Exporting dataset into CSV format file.

write.csv(C1survey, "C1survey_clea.csv", row.names = FALSE)

```
